Data Analysis Methods for the Detection of the Epoch of Reionization

### 2017-08-24 PAPER.tex
status of paper before going on hold for qual studying

The major challenge that faces all 21 cm experiments is isolating a small signal that is buried underneath foregrounds and instrumental systematics that are, when combined, 4-5 orders of magnitude brighter. A clean measurement therefore requires an intimate understanding of the instrument and a rigorous study of data analysis choices. With continual progress being made in the field and HERA on the horizon, it is becoming increasingly important to understand how the methods we choose interact with each other to affect power spectrum results. More specifically, it is imperative to develop techniques and tests that ensure the accuracy and reliability of a potential EoR detection. In this paper, we discuss three themes that are essential to investigate for a robust $21$ cm power spectrum analysis. We also highlight four power spectrum techniques and their trade-offs, traps, and connections to the themes . We first approach the themes from a broad perspective, and then perform a detailed case study using data from the 64-element configuration of PAPER.

This paper is organized as follows. In section 2 we introduce the three themes of our focus, using a toy model to develop intuition for each one. Section 3 presents a case study into each theme using data from the PAPER-64 array, highlighting key changes from \citet{ali_et_al2015}. We conclude in Section 7.

@@

There are analysis techniques to mitigate the effects of foreground leakage and prevent information from low $k's$ from spreading to high $k$ values. For example, narrow window functions can be used to minimize the covariance of a particular $k$ value with other ones (\citealt{liu_et_al2014b}). \cc{math!} In other words, one can construct an estimator using OQE that forces a window function to have a minimum response to low $k$ values. The PAPER-64 window function is constructed in such a way, specifically to prevent foregrounds that live at low $k's$ from contaminating higher $k$-modes (see Section \ref{sec:Bias}). Minimizing foreground leakage in this way however, comes with the trade-off of compromising power spectrum sensitivity, since narrow window functions increases errors for each $k$-mode \cc{develop more or cite?}.

@@

As a brief review, PAPER is a dedicated $21$ cm experiment located in the Karoo Desert in South Africa. The PAPER-64 configuration consists of 64 dual-polarization drift-scan elements (Figure \ref{fig:paper}) that are arranged in a grid layout. For our case study, we focus solely on Stokes I data from PAPER's $30$ m East/West baselines, of which there are a total of $56$. For information about the backend system of PAPER-64, its observations, and data reduction pipeline, we refer the reader to \citet{parsons_et_al2010} and \citet{ali_et_al2015}.

@@

We note that the PAPER-64 dataset that we use in this case study differs from that in \citet{ali_et_al2015} only by the fringe-rate filter. In \citet{ali_et_al2015}, the applied filter was degraded by widening it in fringe-rate space. This was chosen in order to increase the number of independent modes and reduce signal loss. With the development of a robust method for assessing signal loss, we feel comfortable using a narrow filter --- the optimal fringe-rate filter --- in order to maximize sensitivity. This filter is computed for a fiducial $30$ m baseline at $150$ MHz, the center frequency in our band. The filter in both the fringe-rate domain and time domain is shown in Figure \ref{fig:frp}.

@@

We now shift our attention towards computing signal loss for our fringe-rate filtered PAPER-64 dataset.  To recap, we use 30 m East/West baselines of PAPER-64. Our dataset spans $8$ hours of LST ($.5$-$8.6$ hrs), includes a total of $51$ baselines, and is fringe-rate filtered using an optimal fringe-rate filter. We have two datasets (even days and odd days), and only cross-multiply data from different days and different baselines. We are interested in $21$ frequency channels (channels 95-115), which yields a power spectrum for a redshift of $z=8.4$.

While our methodology outlined below is robust to any weighting scheme, here we demonstrate the computation using inverse covariance weighting ($\textbf{R} = \textbf{C}^{-1}$), which leads to substantial loss. With this weighting, our expressions for $P_{in}$ and $P_{out}$ become:

@@

Using the input and output power spectrum estimates for a range of EoR amplitudes, we next describe two methods to determine signal loss associated with the over-fitting of noise from inverse covariance weighting of PAPER-64 data. \cc{describing both methods may be overkill?}

The first method is very straightforward. Post-bootstrapping, our final power spectrum estimates are 1-dimensional and only a function of $k$. For each $k$, we simply look at $P_{out}$ as a function of $P_{in}$, as shown in Figure \ref{fig:sigloss1_data}. The shape of this function can be explained as follows. At small injection levels (small $\textbf{e}$), $P_{out}$ and $P_{in}$ are equal, and there is no signal loss. As the amplitude of EoR increases, we then move into a regime where the final output power spectrum is lower than the unweighted input one. This is dangerous, because without correcting for this effect one might be led to underestimate the EoR signal. The peculiar tail at very low injection levels (where $P_{out} > P_{in}$) is not a physical feature, but rather illustrates that there is some non-negligible power in the cross-terms of Equation \ref{eq:crossterm} when $\textbf{e}$ is very small.

For this method, we interpolate the signal loss factor (per $k$), computed as $P_{in}/P_{out}$, at a $P_{out}$ value equal to the $2\sigma$ power spectrum upper limit of our data. In other words, we look at $\hat{\textbf{p}}_{\alpha} \propto \textbf{x}^{\dagger}\textbf{C}_{x}^{-1}\textbf{Q}_{\alpha}\textbf{C}_{x}^{-1}\textbf{x}$, compute its $2\sigma$ upper limit (mean over bootstraps + $2 \times $standard deviation over bootstraps), and interpolate the value of $P_{in}/P_{out}$ at this value. We therefore end up with one signal loss correction factor per $k$.

@@

Our second method for estimating signal loss is similar to the first, but more comprehensive in a statistical sense. Instead of looking at input and output power spectra after bootstrapping, we now look at their values for every bootstrap in order to get a sense of their distributions. Figure \ref{fig:sigloss2_data} plots $P_{in}$ vs. $P_{out}$ for $20$ bootstraps, and as expected, the function now has a spread in the width-direction in comparison to what was plotted in Figure \ref{fig:sigloss1_data}, but otherwise shows a familiar trend. Similarly, our weighted power spectra also has a defined spread due to bootstrapping.

@@

Using these two distributions ($P_{in}/P_{out}$ and $\hat{\textbf{p}}$), we can create bins along the $\hat{\textbf{p}}$ axis to yield histograms of signal loss factors for each bin. We similarly sort the values of $\hat{\textbf{p}}$ into the same bins, and multiply the probability of $\hat{\textbf{p}}$ per bin (the number of values falling into that bin, divided by the total) with the signal loss factors in that bin, essentially computing a weighted average across all bins to obtain a final signal loss factor per $k$. The signal loss results for this method are very similar to the previous method. For future power spectrum results, we choose to use the second method because it computes final signal loss values using our full distributions of measurements.

\cc{This next paragraph needs to be worked on. What I'm going for is a plot that `proves' our signal loss method is correct, but it's hard to show since the combination of inverse covariance weighting + signal loss correction produces results that are HIGHER than the unweighted level. The plot shown below uses PS points instead of error bars, which brings them into closer agreement...}
To verify our signal loss computations, we compare the unweighted power spectrum estimate of EoR (`true' EoR) with the signal loss corrected weighted power spectrum estimate of EoR + data in a regime where the injected EoR is very large (since there is minimal signal loss for small EoR). We compare these two quantities because we must ensure that our weighted combined dataset, after signal loss correction, does not lie below the level of EoR. From Figure \ref{fig:sigloss_beforeafter}, we see over $3$ orders of magnitude of signal loss associated with the inverse covariance weighting of fringe-rate filtered PAPER-64 data. The power spectrum, post-correction, lies just above the EoR-only power spectrum. However, it exhibits a different shape due to \cc{??}.

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{plots/sigloss_beforeafter.png}
	\caption{Power spectrum estimates of the combined EoR + PAPER-64 dataset, pre (squares) and post (circles) signal loss correction. The dashed blue line is the unweighted power spectrum estimate of EoR alone.}
	\label{fig:sigloss_beforeafter}
\end{figure}

One thing to note is that for both methods, we have been careful to validate that the computations yield no signal loss (signal loss factors of $1$) for the unweighted power spectrum case, as is expected. This is important in confirming that signal loss is a direct result of the choice for $\textbf{C}$. Similarly, we can compute signal loss for intermediate cases as well, such as a regularized $\textbf{C}$ which exhibits some lossiness, but not as much as inverse covariance weighting.

@@

@@

\begin{figure*}
	\centering
	\includegraphics[width=0.4\textwidth]{plots/ps2_data_nosigloss.png}
	\includegraphics[width=0.4\textwidth]{plots/ps2_data.png}
	\caption{Full inverse covariance weighted power spectrum of PAPER-64 data (black points, with $2\sigma$ error bars) before signal loss correction (left) and after (right). The dashed blue line is the unweighted power spectrum ($2\sigma$ upper limit). The solid green line is the theoretical noise level prediction based on observational parameters.}
	\label{fig:ps2_data}
\end{figure*}

Following our toy model investigation in Section \ref{sec:SiglossOverview}, we investigate the shape of the eigenspectrum of $\textbf{C}$ for a typical baseline used in the analysis. Figure \ref{fig:eigenspectrum} shows this spectrum for baseline (1,4). The spectrum is very steep, spanning 4 orders of magnitude, implying that a couple modes are weighted much more than the rest. In fact, this behavior is much more extreme than we saw in the toy model, whose eigenspectrum (for the `fringe-rate filtered' case) only started to vividly steepen for the last few eigenmodes. The consequence of assigning individual weights to many more modes than are present in the data is significant --- to the point of resulting in power spectrum values that are higher than if no weighting scheme was used at all.

\begin{figure*}
	\centering
	\includegraphics[width=0.4\textwidth]{plots/eigenspectrum.png}
	\includegraphics[width=0.4\textwidth]{plots/covariance.png}
	\caption{Eigenspectrum for $\textbf{C}$ for baseline (1,4) for the 21 channels ofs interest (left) and covariance matrix $\textbf{C}$ for the same baseline (right).}
	\label{fig:eigenspectrum}
\end{figure*}

@@

@@

>>>>

The major challenge that faces all 21 cm experiments is isolating a small signal that is buried underneath foregrounds and instrumental systematics that are, when combined, 4-5 orders of magnitude brighter. A clean measurement therefore requires an intimate understanding of the instrument and a rigorous study of data analysis choices. With continual progress being made in the field and HERA on the horizon, it is becoming increasingly important to understand how the methods we choose interact with each other to affect power spectrum results. More specifically, it is imperative to develop techniques and tests that ensure the accuracy and reliability of a potential EoR detection. In this paper, we discuss three themes that are essential to investigate for a robust $21$ cm power spectrum analysis. We also highlight four power spectrum techniques and their trade-offs, traps, and connections to the themes. We first approach the themes from a broad perspective, and then perform a detailed case study using data from the 64-element configuration of PAPER.

This paper is organized as follows. In section 2 we introduce the three themes of our focus, using a toy model to develop intuition for each one. Section 3 presents a case study into each theme using data from the PAPER-64 array, highlighting key changes from \citet{ali_et_al2015} which have led to a revised PAPER-64 power spectrum limit. We conclude in Section 4.

@@

There are analysis techniques to mitigate the effects of foreground leakage and prevent information from low $k's$ from spreading to high $k$ values. For example, narrow window functions can be used to minimize the covariance of a particular $k$ value with other ones (\citealt{liu_et_al2014b}). In other words, one can construct an estimator using OQE that forces a window function to have a minimum response to low $k$ values. The PAPER-64 window function is constructed in such a way, specifically to prevent foregrounds that live at low $k's$ from contaminating higher $k$-modes (see Section \ref{sec:Bias}). Minimizing foreground leakage in this way however, comes with the trade-off of compromising power spectrum sensitivity, since narrow window functions increases errors for each $k$-mode \cc{develop more or cite?}.

@@

As a brief review, PAPER is a dedicated $21$ cm experiment located in the Karoo Desert in South Africa. The PAPER-64 configuration consists of 64 dual-polarization drift-scan elements (Figure \ref{fig:paper}) that are arranged in a grid layout. For our case study, we focus solely on Stokes I data from PAPER's $30$ m East/West baselines. For information about the backend system of PAPER-64, its observations, and data reduction pipeline, we refer the reader to \citet{parsons_et_al2010} and \citet{ali_et_al2015}.

@@

For our analysis, we use $8$ hours of LST (RA $0.5$-$8.6$ hours) and $51$ total baselines. All power spectrum results are produced using channels $95$-$115$, corresponding to a center redshift of $z=8.4$. We note that the PAPER-64 dataset that we use in this case study differs from that in \citet{ali_et_al2015} mainly by the fringe-rate filter. In \citet{ali_et_al2015}, the applied filter was degraded by widening it in fringe-rate space. This was chosen in order to increase the number of independent modes and reduce signal loss. With the development of a robust method for assessing signal loss, we feel comfortable using a narrow filter --- the optimal fringe-rate filter --- in order to maximize sensitivity. This filter is computed for a fiducial $30$ m baseline at $150$ MHz, the center frequency in our band. The filter in both the fringe-rate domain and time domain is shown in Figure \ref{fig:frp}.

@@

\cc{This section needs to be filled in once our method is finalized...}

We now shift our attention towards computing signal loss for our fringe-rate filtered PAPER-64 dataset. While our methodology outlined below is robust to any weighting scheme, here we demonstrate the computation using inverse covariance weighting ($\textbf{R} = \textbf{C}^{-1}$), which leads to substantial loss. With this weighting, our expressions for $P_{in}$ and $P_{out}$ become:

@@

@@

@@

%\begin{figure}
%	\centering
%	\includegraphics[width=\columnwidth]{plots/sigloss_beforeafter.png}
%	\caption{Power spectrum estimates of the combined EoR + PAPER-64 dataset, pre (squares) and post (circles) signal loss correction. The dashed blue line is the unweighted power spectrum estimate of EoR alone.}
%	\label{fig:sigloss_beforeafter}
%\end{figure}

@@

\cc{This section is under construction...}

@@

%\begin{figure*}
%	\centering
%	\includegraphics[width=0.4\textwidth]{plots/ps2_data_nosigloss.png}
%	\includegraphics[width=0.4\textwidth]{plots/ps2_data.png}
%	\caption{Full inverse covariance weighted power spectrum of PAPER-64 data (black points, with $2\sigma$ error bars) before signal loss correction (left) and after (right). The dashed blue line is the unweighted power spectrum ($2\sigma$ upper limit). The solid green line is the theoretical noise level prediction based on observational parameters.}
%	\label{fig:ps2_data}
%\end{figure*}

Following our toy model investigation in Section \ref{sec:SiglossOverview}, we investigate the shape of the eigenspectrum of $\textbf{C}$ for a typical baseline used in the analysis. Figure \ref{fig:eigenspectrum} shows this spectrum for baseline (1,4). The spectrum is very steep, spanning 4 orders of magnitude, implying that different modes are given dramatically different weights. In fact, this behavior is much more extreme than we saw in the toy model, whose eigenspectrum (for the `fringe-rate filtered' case) only started to vividly steepen for the last few eigenmodes. The consequence of assigning individual weights to many more modes than are present in the data is significant --- to the point of resulting in power spectrum values that are higher than if no weighting scheme was used at all.

%\begin{figure*}
%	\centering
%	\includegraphics[width=0.4\textwidth]{plots/eigenspectrum.png}
%	\includegraphics[width=0.4\textwidth]{plots/covariance.png}
%	\caption{Eigenspectrum for $\textbf{C}$ for baseline (1,4) for the 21 channels ofs interest (left) and covariance matrix $\textbf{C}$ for the same baseline (right).}
%	\label{fig:eigenspectrum}
%\end{figure*}

@@

\begin{figure*}
	\centering
	\includegraphics[width=0.4\textwidth]{plots/ps1_data_nosigloss.png}
	\includegraphics[width=0.4\textwidth]{plots/ps2_data_nosigloss.png}
	\caption{Highest sensitivity power spectrum of PAPER-64 using $\hat{\textbf{C}}_{eff} = \hat{\textbf{C}} * \textbf{I}$ (minimal signal loss). Black and grey points correspond to positive and negative power spectrum values, respectively, with $2\sigma$ error bars also plotted. The dashed blue line is the unweighted power spectrum ($2\sigma$ upper limit). The solid green line is the theoretical noise level prediction based on observational parameters. This power spectrum result differs from \citet{ali_et_al2015} in that it only uses data from one type of baseline ($30$ m East/West baselines) instead of three. Additionally, an optimal fringe-rate filter was used instead of a degraded one. Major differences from previously published results stem from revisions regarding signal loss, bootstrapping, and the theoretical error computation.}
	\label{fig:ps1_data}
\end{figure*}

@@

The highest sensitivity power spectrum results using PAPER-64 data, shown in Figure \ref{fig:ps1_data}, have positive biases. As discussed in Section \ref{sec:BiasTypes}, the detections that appear at low $k$ values are most likely attributable to foreground leakage. However, the positive biases at higher $k$ values require further investigation.
